A Lanczos Procedure for Approximating Eigenvalues of Large Stochastic Matrices

### 2013-10-18 tex/InvariantSubspaces.tex
minor mods

\section{Approximate Invariant Subspace}
% Sec 3.2
\label{sec:appr-invar-subsp}
%% We qualified the foregoing by stating that the facts are only theoretically
%% useful. This
We noted above that Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are
\emph{theoretically} useful. Speaking practically, however, 
%% This is because 
when choosing observables on the state space we may not be sure that
they will satisfy the 
%
%
% -------------16.txt---------------------------------------------------
%
%
primary assumption underlying these facts. %Facts 3.1.1 and 3.1.2. 
Recall the assumption: $\ran(\Psi)$ is an invariant
subspace for $\M$ where $\Psi = \Pi^{1/2}(\Phi - \<\Phi\>)$. It may well be the
case that there is a $\psi_j\in \ran(\Psi)$ such that 
$M\psi_j \notin \ran(\Psi)$, thereby violating the assumption. 
However, lacking an invariant subspace, we 
might hope that our observables at least provide an \emph{approximate invariant subspace}.
\begin{definition}
%% Definition 3.2.1
If the columns of $\Psi\in \R^{d\times p}$ are independent and the 
\emph{residual matrix} $\E = \M\Psi - \Psi \T$ is small for some 
$\T \in \R^{p\times p}$, then
$\ran(\Psi)$ defines an \emph{approximate invariant subspace}. 
\end{definition}
To see how an approximate invariant subspace can be useful for approximating
eigenvalues of $\M$, we recall a theorem from Golub and Van Loan~\cite{Golub:1996}.
\begin{theorem}
% Theorem 3.2.1
\label{thm:3.2.1}  
Suppose $\M \in R^{d\times d}$ and $T \in \R^{p\times p}$ are symmetric and let
$E = \M\Q - \Q\T$, where $\Q \in \R^{d\times p}$ is orthonormal 
(i.e., $\Q^t\Q = \I$). Then there exist $\mu_1,\dots, \mu_p \in \lambda(\T)$ and
$\lambda_1, \dots, \lambda_p \in \lambda(\M)$ such that
\[
|\mu_k - \lambda_k| \leq \sqrt{2}\|\E\|_2, \quad \text{ for $k = 1, \dots, p$.}
\]
\end{theorem}
If the subspace $\ran(\Q)$ is an approximate invariant subspace, then the definition implies that
there is a choice $\T$ rendering the error $\|\E\|_2$ small, and thus the
eigenvalues of $\T$ provide a good approximation to those of $\M$.

@@

The second equality in Theorem~\ref{thm:3.2.2} is a consequence of the polar decomposition, in
which $\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2}$. Therefore,
$\T = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}$
is the minimizer, as claimed.
\end{proof}

@@

Now, $\rho(\psi_1)$ changes most rapidly in the direction of the gradient
$\nabla \rho(\psi_1)$.
\begin{equation}
  \label{eq:3.4}
\nabla \rho(\psi_1) = \left(\frac{\partial \rho(\psi_1)}{\partial \psi_1(1)}, \dots,
\frac{\partial \rho(\psi_1)}{\partial \psi_1(d)}\right)
 = \frac{2}{\psi_1^t\psi_1} (\M\psi_1 - \rho(\psi_1)\psi_1).
\end{equation}
So, to maximize the left hand side of~(\ref{eq:3.3}), $\Psi_2$ should be chosen
so that the subspace $\ran(\Psi_2)$ contains the gradient vector.
That is, we must choose $\psi_2$ so that
\begin{equation}
\label{eq:3.5}
\nabla \rho(\psi_1) \in \ran\{\psi_1, \psi_2\} = \ran(\Psi_2).
\end{equation}
Clearly, Equation~(\ref{eq:3.4}) implies 
$\nabla \rho(\psi_1) \in \ran\{\psi_1, \M\psi_1\}$.
Therefore, if $\ran\{\psi_1, \psi_2\} = \ran\{\psi_1, \M\psi_1\}$, 
then~(\ref{eq:3.5}) is satisfied.
%
%
% ------------ 21.txt ----------------------------------------------
%
%
In general, having chosen $\Psi_k = [\psi_1 \; \cdots \; \psi_k]$ so that
\[
\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
\M\psi_k\},
\]
we must chose $\psi_{k+1}$ so that for any nonzero vector $a\in \R^k$,
\begin{equation}
\label{eq:3.6}
\nabla \rho(\Psi_ka) \in \ran\{\psi_1, \psi_2, \dots, \psi_k\}.
\end{equation}
Now,
\[
\nabla \rho(\Psi_ka)  = 
 = \frac{2}{a^t\Psi_k^t\Psi_ka} (\M\Psi_k - \rho(\Psi_ka)\Psi_ka).
\]
and therefore,
\[
\nabla \rho(\Psi_ka)  \in 
\ran(\M\Psi_k) \cup \ran(\Psi_k) = \ran\{\psi_1, \M\psi_1, \dots, \M^k\psi_1\}.
\]
Thus, the requirement~(\ref{eq:3.6}) is satisfied when 
$\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots, \M^k\psi_1\}$.

>>>>

\section{Approximate Invariant Subspace}
% Sec 3.2
\label{sec:appr-invar-subsp}
%% We qualified the foregoing by stating that the facts are only theoretically
%% useful. This
We noted above that Facts~\ref{fact:3.1.1} and~\ref{fact:3.1.2} are
theoretically useful. Speaking practically now,
%% This is because 
when choosing observables on the state space we may not be sure that
they will satisfy the 
%
%
% -------------16.txt---------------------------------------------------
%
%
primary assumption underlying the two facts. %Facts 3.1.1 and 3.1.2. 
Recall the assumption: $\ran(\Psi)$ is an invariant
subspace for $\M$ where $\Psi = \Pi^{1/2}(\Phi - \<\Phi\>)$. It may well be the
case that there exists $\psi_j\in \ran(\Psi)$ such that 
$\M\psi_j \notin \ran(\Psi)$, thereby violating the assumption. 
Even if we are lacking an invariant subspace, however, for some applications it
is reasonable to expect that observables can be chosen to provide at least
an \emph{approximate invariant subspace}, which is defined as follows:
\begin{definition}
%% Definition 3.2.1
If the columns of $\Psi\in \R^{d\times p}$ are independent and the norm of the
\emph{residual matrix} $\E = \M\Psi - \Psi \T$ is small for some  
$\T \in \R^{p\times p}$, then
$\ran(\Psi)$ defines an \emph{approximate invariant subspace}. 
\end{definition}
To see how an approximate invariant subspace can be useful for approximating
eigenvalues of $\M$, we recall a theorem from Golub and Van Loan~\cite{Golub:1996}.
\begin{theorem}
% Theorem 3.2.1
\label{thm:3.2.1}  
Suppose $\M \in R^{d\times d}$ and $\T \in \R^{p\times p}$ are symmetric and let
$\E = \M\Q - \Q\T$, where $\Q \in \R^{d\times p}$ is orthonormal 
(i.e., $\Q^t\Q = \I$). Then there exist $\mu_1,\dots, \mu_p \in \lambda(\T)$ and
$\lambda_1, \dots, \lambda_p \in \lambda(\M)$ such that
\[
|\mu_k - \lambda_k| \leq \sqrt{2}\|\E\|_2, \quad \text{ for $k = 1, \dots, p$.}
\]
\end{theorem}
If the subspace $\ran(\Q)$ is an approximate invariant subspace, then the definition implies that
there is a choice $\T$ rendering the error $\|\E\|_2$ small, and thus the
eigenvalues of $\T$ provide a good approximation to those of $\M$.

@@

The second equality in Theorem~\ref{thm:3.2.2} is a consequence of the polar decomposition, in
which $\Q = \Psi \Z^{-1} = \Psi(\Psi^t\Psi)^{-1/2}$. Thus,
$\T = (\Psi^t\Psi)^{-1/2}\Psi^t \M^t \Psi (\Psi^t\Psi)^{-1/2}$
is the minimizer, as claimed.
\end{proof}

@@

Now, $\rho(\psi_1)$ changes most rapidly in the direction of the gradient
$\nabla \rho(\psi_1)$.
\begin{equation}
  \label{eq:3.4}
\nabla \rho(\psi_1) = \left(\frac{\partial \rho(\psi_1)}{\partial \psi_1(1)}, \dots,
\frac{\partial \rho(\psi_1)}{\partial \psi_1(d)}\right)
 = \frac{2}{\psi_1^t\psi_1} (\M\psi_1 - \rho(\psi_1)\psi_1).
\end{equation}
So, to maximize the left hand side of~(\ref{eq:3.3}), $\Psi_2$ should be chosen
so that the subspace $\ran(\Psi_2)$ contains the gradient vector.
That is, we must choose $\psi_2$ so that
\begin{equation}
\label{eq:3.5}
\nabla \rho(\psi_1) \in \ran\{\psi_1, \psi_2\} = \ran(\Psi_2).
\end{equation}
Clearly, Equation~(\ref{eq:3.4}) implies 
$\nabla \rho(\psi_1) \in \ran\{\psi_1, \M\psi_1\}$.
Thus, if $\ran\{\psi_1, \psi_2\} = \ran\{\psi_1, \M\psi_1\}$, 
then~(\ref{eq:3.5}) is satisfied.
%
%
% ------------ 21.txt ----------------------------------------------
%
%
In general, having chosen $\Psi_k = [\psi_1 \; \cdots \; \psi_k]$ so that
$\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
\M\psi_k\}$,
we must chose $\psi_{k+1}$ so that for any nonzero vector $a\in \R^k$,
\begin{equation}
\label{eq:3.6}
\nabla \rho(\Psi_ka) \in \ran\{\psi_1, \psi_2, \dots, \psi_k\}.
\end{equation}
Now,
\[
\nabla \rho(\Psi_ka)  = 
 = \frac{2}{a^t\Psi_k^t\Psi_ka} (\M\Psi_k - \rho(\Psi_ka)\Psi_ka).
\]
and therefore,
\[
\nabla \rho(\Psi_ka)  \in 
\ran(\M\Psi_k) \cup \ran(\Psi_k) = \ran\{\psi_1, \M\psi_1, \dots, \M^k\psi_1\}.
\]
Thus, the requirement~(\ref{eq:3.6}) is satisfied when 
\[
\ran\{\psi_1, \psi_2, \dots, \psi_k\} = \ran\{\psi_1, \M\psi_1, \dots,
\M^k\psi_1\}.
\]
